---
title: "Handling Missing Values"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
tutorial:
  id: "Missing Values"
  version: 0.5
---


```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(datasets)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(exercise.checker = gradethis::grade_learnr)
learnr::tutorial_options(
  exercise.timelimit = 60)
```

## Missing Values

###

One of the common tasks in data analysis is handling missing values.

In R, missing values are often represented by the symbol `NA` (not available) or some other value that represents missing values (i.e. `99`).

Impossible values (e.g., dividing by zero) are represented by the symbol `NaN` (not a number).

Working with missing data in R is not very hard. In this tutorial you will learn how to

 * Test for Missing Values
 * Recode Missing Values
 * Exclude Missing Values
 
## Test for Missing Values

###

To identify missing values you can use `is.na()` which returns a logical vector with `TRUE` in the element locations that contain missing values represented by `NA`. `is.na()` will work on vectors, lists, matrices, and data frames.

```{r  echo = TRUE}
# vector with missing data
x <- c(1:4, NA, 6:7, NA)
x

# checking if x contains missing data
is.na(x)

# data frame with missing data
df <- data.frame(col1 = c(1:3, NA),
                 col2 = c("this", NA,"is", "text"), 
                 col3 = c(TRUE, FALSE, TRUE, TRUE), 
                 col4 = c(2.5, 4.2, 3.2, NA),
                 stringsAsFactors = FALSE)

# identify NAs in full data frame
is.na(df)


# identify NAs in specific data frame column
is.na(df$col4)
```

To identify the location or the number of `NA`-s we can leverage the `which()` and `sum()` functions:

```{r echo=TRUE}
# identify location of NAs in vector
which(is.na(x))


# identify count of NAs in data frame
sum(is.na(df))
```

For data frames, a convenient shortcut to compute the total missing values in each column is to use `colSums()`:

```{r echo=TRUE}
colSums(is.na(df))
```

## Recode Missing Values

###

To recode missing values; or recode specific indicators that represent missing values, you can use normal subsetting and assignment operations. For example, you can recode missing values in vector `x` with the mean values in `x` by first subsetting the vector to identify `NA`s and then assign these elements a value. Similarly, if missing values are represented by another value (i.e. `99`) you can simply subset the data for the elements that contain that value and then assign a desired value to those elements. Let's look at the concepts using some simple examples.

```{r echo=TRUE}
# recode missing values with mean
# vector with missing data

x <- c(2:5,NA,7:10,NA,12)
x

x[is.na(x)] <- mean(x, na.rm = TRUE)

# rounding the values in x upto 2 decimal places

round(x,2)

# data frame that codes missing values as 99
df1 <- data.frame(col1 = c(1:5,99), col2 = c(2,5.2,99,3.78,99,9.13))

#  change 99s to NAs

df1[df1 == 99] <- NA
df1
```

If we want to recode missing values in a single data frame variable we can subset for the missing value in that specific variable of interest and then assign it the replacement value. For example, here we recode the missing value in `col4` with the mean value of `col4`.

```{r echo=TRUE}
# data frame with missing data

df2 <- data.frame(col1 = c(1:3,NA),
                  col2 = c("this",NA,"is","text"),
                  col3 = c(TRUE, FALSE, FALSE, TRUE),
                  col4 = c(1.3,2.4,NA,9.5),
                  stringsAsFactors = FALSE)
df2$col4[is.na(df2$col4)] <- mean(df2$col4, na.rm = TRUE)
df2
```

## Exclude Missing Values

### 

We can exclude missing values in a couple different ways. First, if we want to exclude missing values from mathematical operations use the `na.rm = TRUE` argument. If you do not exclude these values most functions will return an `NA`.

```{r echo=TRUE}
# A vector with missing values
x <- c(5:8, NA, 1:3, NA)

# including NA values will produce an NA output
mean_NA <- mean(x)
mean_NA

#  excluding NA values will calculate the mathematical operation for all non-missing values
mean_nonNA <- mean(x, na.rm = TRUE) 
mean_nonNA
```

You may also desire to subset our data to obtain complete observations, those observations (rows) in the data that contain no missing data.You can do this in a few different ways. Considering our previous data frame `df2` here.

First, to find complete cases (rows with no missing values) we can leverage the `complete.cases()` function which returns a logical vector identifying rows which are complete cases.

```{r echo=TRUE}
df2
complete.cases(df2)
```

So, looking back at the data frame we made, `df2` you can see that rows 1 and 3 are complete cases. We can use this information to subset our data frame which will return the rows which `complete.cases()` found to be TRUE.

```{r echo=TRUE}
# subset with complete.cases to get complete cases
df2[complete.cases(df2), ]

# or subset with `!` operator to get incomplete cases
df2[!complete.cases(df), ]
```

A shorthand alternative is to simply use `na.omit()` to omit all rows containing missing values.

```{r echo=TRUE}

# use na.omit() to get rows with no missing values
na.omit(df2)
```


## Exercises

###

Consider the `airquality` data set here in package `datasets`.

```{r echo=TRUE}
data("airquality")
```

### Exercise 1

Write down the piece of code you would use to find how many missing values are there corresponding to each column in the `airquality` dataset.

```{r exercise1, exercise=TRUE}
```

```{r exercise1-check}
grade_result(
  pass_if(~ identical(.result, colSums(is.na(airquality))), "Good Job!"),
  fail_if(~ TRUE)
)
```

### Exercise 2

How would you impute the the missing values with their median?

```{r exercise2, exercise=TRUE}
```

```{r exercise2-solution}
airquality$Ozone[is.na(airquality$Ozone)] = median(airquality$Ozone,na.rm = TRUE)
airquality$Solar.R[is.na(airquality$Solar.R)] = median(airquality$Solar.R, na.rm = TRUE)
```

```{r exercise2-check}
grade_code("Excellent")
```

### Exercise 3

How would you omit all rows containing missing values?

```{r exercise3, exercise=TRUE}
```

```{r exercise3-check}
grade_result(
  pass_if(~ identical(.result, na.omit(airquality)), "Well done!"),
  fail_if(~ TRUE)
)
```

### Submitting Work

```{r submission-setup}
submission_code <- function(UID){
  httr::sha1_hash('missval1',as.character(UID))
}
```

Generate your submission code by putting in your UID in the function below. For example, if your UID is `2`, then your code should look like `submission_code(UID = 2)`
```{r submission, exercise = TRUE}
# Replace the number below with your UID
submission_code(2)
```


