---
title: "String Manipulation with Base R"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
tutorial:
  id: "StringBaseR"
  version: 0.5
---


```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(datasets)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(exercise.checker = gradethis::grade_learnr)
learnr::tutorial_options(
  exercise.timelimit = 60)
```

## String Manipulation

###

Basic string manipulation typically includes case conversion, simple character replacement, abbreviating, substring replacement, adding/removing white space, and performing set operations to compare similarities and differences between two character vectors.These operations can all be performed with base R functions. 

This tutorial illustrates base R string manipulation for `case conversion`, `simple character replacement`, `abbreviating`, and `substring replacement`. 

However, there are other powerful packages like `stringr` in R to perform the tasks and some more advanced things as well.

## Case conversion

###

To convert all upper case characters to lower case use `tolower()`:

```{r echo=TRUE}
x <- "Learning To MANIPULATE strinGS in R"

tolower(x)
```

To convert all lower case characters to upper case use `toupper()`:

```{r echo= TRUE}
toupper(x)
```

## Simple Character Replacement

###

To replace a character (or multiple characters) in a string you can use `chartr()`:

```{r echo=TRUE}
# replace 'A' with 'a'
x <- "This is A string."
chartr(old = "A", new = "a", x)

# multiple character replacements
# replace any 'd' with 't' and any 'z' with 'a'
y <- "Tomorrow I plzn do lezrn zbout dexduzl znzlysis."
chartr(old = "dz", new = "ta", y)
```

Note that `chartr()` replaces every identified letter for replacement so the only time you use it is when you are certain that you want to change every possible occurrence of a letter.

## String Abbreviations

###

To abbreviate strings you can use `abbreviate()`:

```{r echo=TRUE}
states <- c("Maryland","Pennsylvania","California","Michigan","Wisconsin")

# default abbreviations
abbreviate(states)

# set minimum length of abbreviation
abbreviate(states, minlength = 2)
```

**Note** 

If you are working with U.S. states, R already has a pre-built vector with state names `state.name`. Also, there is a pre-built vector of abbreviated state names `state.abb`.

## Extract/Replace Substrings

###

To extract or replace substrings in a character vector there are three primary base R functions to use: `substr()`, `substring()`, and `strsplit()`.

The purpose of `substr()` is to extract and replace substrings with specified starting and stopping characters:

```{r echo=TRUE}

string <- c("stringmanipulationisfun")

# extract 5th character in string
substr(string, start = 5, stop = 5)


# extract 5th-9th characters in string
substr(string, start = 5, stop = 9)


# replace 5th-9th characters with `R`
substr(string, start = 5, stop = 9) <- "RRRRRR"
string
```

The purpose of `substring()` is to extract and replace substrings with only a specified starting point. `substring()` also allows you to extract/replace in a recursive fashion:

```{r echo=TRUE}
alphabet <- paste(LETTERS, collapse = "")

# extract 18th through last character
substring(alphabet, first = 18)


# recursive extraction; specify start position only
substring(alphabet, first = 18:24)


# recursive extraction; specify start and stop positions
substring(alphabet, first = 1:5, last = 3:7)
```

To split the elements of a character string use `strsplit()`:

```{r echo=TRUE}
z <- "The day after let us take a break and watch a movie."
strsplit(z, split = " ")

a <- "Alabama-Alaska-Arizona-Arkansas-California"
strsplit(a, split = "-")
```

Note that the output of `strsplit()` is a list. To convert the output to a simple atomic vector simply wrap in `unlist()`.

```{r echo=TRUE}
unlist(strsplit(a, split = "-"))
```

## Exercises

### Exercise 1 

Consider the following sentence on `Gross Domestic Product`as a string object:

"GDP is a monetary MEASURE of the market value of all the final goods and services produced and sold (not resold) in a specific time PERIOD by countries."

Store it in a string object named `string`.Convert the lower case letters in this sentence to upper cases. Write down the code below.

```{r exercise1, exercise=TRUE}
```

```{r exercise1-solution}
string <- c("gdp is a monetary measure of the market value of all the final goods and services produced and sold (not resold) in a specific time period by countries.")

toupper(string)
```

```{r exercise1-check}
grade_code("Excellent")
```

### Exercise 2

Create a string object named `streets` to store the names of the streets : Rochester, Amsterdam, Temple, Wellington, Boulevard, Regents, Lewisdale. Abbreviate the names upto 3 letters. Write down the necessary code for the purpose in the following code chunk. 

```{r exercise2, exercise=TRUE}
```


```{r exercise2-solution}
streets <- c("Rochester","Amsterdam","Temple","Wellington","Boulevard","Regents","Lewisdale")
abbreviate(streets, minlength = 3)
```

```{r exercise2-check}
grade_code("Excellent")
```

### Exercise 3

Consider the word `XXXnksgiving`, store this as a character string in R with the name `string`. Replace `XXX` with `tha` and display the new string. Write down the pertinent code below.

```{r exercise3, exercise=TRUE}
```

```{r exercise3-solution}
string <- c("XXXnksgiving")
substr(string, start = 1, stop = 3) <- "tha"
string
```

```{r exercise3-check}
grade_code("Excellent!")
```

### Exercise 4

Create a string object named `text` to store the following sentence :

**Older adults and people who have severe underlying medical conditions like heart or lung disease or diabetes seem to be at higher risk for developing more serious complications from COVID-19 illness**

Split the string into individual words and display them as an atomic vector.

```{r exercise4, exercise=TRUE}
```

```{r exercise4-solution}

text <- c("Older adults and people who have severe underlying medical conditions like heart or lung disease or diabetes seem to be at higher risk for developing more serious complications from COVID-19 illness")
unlist(strsplit(text, split = " "))
```

```{r exercise4-check}
grade_code("Excellent!")
```

### Submitting Work

```{r submission-setup}
submission_code <- function(UID){
  httr::sha1_hash('string1',as.character(UID))
}
```

Generate your submission code by putting in your UID in the function below. For example, if your UID is `2`, then your code should look like `submission_code(UID = 2)`
```{r submission, exercise = TRUE}
# Replace the number below with your UID
submission_code(2)
```


